---
title: "Archaeological bone decontamination"
author: "Zandra Fagernäs"
date: "17.10.2024"
---

This R Notebook contains code for the bone decontamination portion of "Cleaning the Dead: Optimized decontamination enhances palaeoproteomic analyses of a Pleistocene hominin tooth from Khudji, Tajikistan" by Fagernäs et al.

The metadata file is in supplementary information of the publication, called supplementary_data_1.txt. All analyses are based on MaxQuant output, and can be found on the ProteomeXchange Consortium via the PRIDE partner repository with the dataset identifier PXD050393. Note that there are three separate MaxQuant runs, the folders are indicated in the code below. To run the code, you need to download the files from PRIDE, unzip them, and move them into the same directory as where this Markdown file is saved. Note that  the figures may have been manually edited for improved visuals after exporting. Not all figures that are coded for here are included in the manuscript (comments in the code will show which ones are), as I like to visualize data just to understand it.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(MetBrewer)
library(janitor)
library(ggpubr)
library(Biostrings)
library(car)
library(tidyverse)

meta <- read_delim("supplementary_data_1.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  select(-species)
```

## Protein number

```{r}
# Load dataset
proteins <- read_delim("equus_canis_maxquant/proteinGroups.txt", 
    delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(str_detect(protein_i_ds, "HORSE|CANLF")) %>%
  select(protein_i_ds, starts_with("razor_unique_peptides_")) 

# Total number of proteins and peptides
protein_total <- proteins %>%
  pivot_longer(names_to = "stagetip", values_to = "peptides", 2:23) %>%
  mutate(stagetip = gsub('razor_unique_peptides_', '', stagetip)) %>%
  filter(peptides > 0) %>%
  group_by(stagetip) %>%
  summarize(protein_count = n(), 
            peptide_count = sum(peptides))

# Add metadata
protein_total$stagetip <- toupper(protein_total$stagetip)
protein_total <- left_join(protein_total, meta)

# Summarize
protein_summary <- protein_total %>%
  group_by(method) %>%
  summarize(mean_protein = mean(protein_count), sd_protein = sd(protein_count), 
            mean_peptide = mean(peptide_count), sd_peptide = sd(peptide_count))

# Assign if proteins are dog or horse, remove ones assigned to both
proteins <- proteins %>%
  mutate(dog = str_detect(protein_i_ds, "CANLF")) %>%
  mutate(horse = str_detect(protein_i_ds, "HORSE"))  %>%
  mutate(species = if_else(dog == TRUE & horse == FALSE, "dog", 
                           if_else(horse == TRUE & dog == FALSE, "horse", "both"))) %>%
  select(-horse, -dog) %>%
  filter(!species == "both")

# Protein count
protein_count <- proteins %>%
  select(species, protein_i_ds, 2:24) %>%
  pivot_longer(names_to = "stagetip", values_to = "peptides", 3:24) %>%
  mutate(stagetip = gsub('razor_unique_peptides_', '', stagetip)) %>%
  filter(peptides > 0) %>%
  group_by(stagetip, species) %>%
  summarize(protein_count = n())

# Add metadata
protein_count$stagetip <- toupper(protein_count$stagetip)
protein_count <- left_join(protein_count, meta)

# Protein mean count
protein_count_mean <- protein_count %>%
  group_by(method, species) %>%
  summarize(mean_protein = mean(protein_count), sd_protein = sd(protein_count))

# Ratio dog:horse protein
protein_ratio <- protein_count %>%
  pivot_wider(values_from = "protein_count", names_from = "species") %>%
  mutate(ratio = dog/horse)
protein_ratio_summary <- protein_ratio %>%
  group_by(method) %>%
  summarize(mean_ratio = mean(ratio), sd_ratio = sd(ratio))

# ANOVA
aov_test <- aov(ratio ~ method, data = protein_ratio)

# Summary of the analysis
summary(aov_test)

# Post-hoc test
TukeyHSD(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(ratio ~ method, data = protein_ratio)

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals)

# Summarize
protein_summary <- protein_count %>%
  group_by(species, method) %>%
  summarize(mean_number = mean(protein_count), sd_number = sd(protein_count))

# Change labels
protein_count_mean <- protein_count_mean %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))
protein_count_mean <- protein_count_mean %>%
    mutate(species = gsub('dog', 'Dog', species)) %>%
    mutate(species = gsub('horse', 'Horse', species))

# Reorder for plotting
protein_count_mean$method <- factor(protein_count_mean$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot (produces Figure 3A)
protein_count_fig <- ggplot(protein_count_mean, aes(x=method, y=mean_protein, fill=species)) + 
  geom_bar(stat="identity", position ="dodge") +
  geom_errorbar(aes(ymin=mean_protein-sd_protein, ymax=mean_protein+sd_protein), 
                width=0.2, position=position_dodge(0.9)) +
  scale_fill_manual(values=met.brewer("Cassatt2", 3), name="Species") +
  xlab("") +
  ylab("Protein number") +
  theme_bw()
```

## PSM number

```{r}
# Load dataset
peptide <- read_delim("equus_canis_maxquant/evidence.txt", 
    delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  #filter(str_detect(proteins, "HORSE|CANLF")) %>%
  filter(intensity > 0) %>%
  select(proteins, experiment, intensity)

# Assign if proteins are dog or horse, remove ones assigned to both
peptide <- peptide %>%
  mutate(dog = str_detect(proteins, "CANLF")) %>%
  mutate(horse = str_detect(proteins, "HORSE")) %>%
  mutate(species = if_else(dog == TRUE & horse == FALSE, "dog", 
                           if_else(horse == TRUE & dog == FALSE, "horse", "both"))) %>%
  select(-horse, -dog) %>%
  filter(!species == "both")

# Peptide count
peptide_count <- peptide %>%
  group_by(experiment, species) %>%
  summarize(peptide_count = n())

# Ratio dog:horse peptide
peptide_ratio <- peptide_count %>%
  pivot_wider(values_from = "peptide_count", names_from = "species") %>%
  mutate(ratio = dog/horse)
peptide_ratio <- left_join(peptide_ratio, meta, by=c("experiment"="stagetip"))

# Summary
ratio_summary <- peptide_ratio %>%
  group_by(method) %>%
  summarize(mean_ratio = mean(ratio), sd_ratio = sd(ratio))

# ANOVA
aov_test <- aov(ratio ~ method, data = peptide_ratio)

# Summary of the analysis
summary(aov_test)

# Post-hoc test
TukeyHSD(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(ratio ~ method, data = peptide_ratio_stats)

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals)

# Add metadata
peptide_count <- left_join(peptide_count, meta, by=c("experiment"="stagetip"))

# Peptide mean count
peptide_count_mean <- peptide_count %>%
  group_by(method, species) %>%
  summarize(mean_peptide = mean(peptide_count), sd_peptide = sd(peptide_count))

# Change labels
peptide_count_mean <- peptide_count_mean %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))
peptide_count_mean <- peptide_count_mean %>%
    mutate(species = gsub('dog', 'Dog', species)) %>%
    mutate(species = gsub('horse', 'Horse', species))

# Reorder for plotting
peptide_count_mean$method <- factor(peptide_count_mean$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot (produces Figure 3B)
peptide_fig <- ggplot(peptide_count_mean, aes(x=method, y=mean_peptide, fill=species)) + 
  geom_bar(stat="identity", position ="dodge") +
  geom_errorbar(aes(ymin=mean_peptide-sd_peptide, ymax=mean_peptide+sd_peptide), 
                width=0.2, position=position_dodge(0.9)) +
  scale_fill_manual(values=met.brewer("Cassatt2", 3), name="Species") +
  xlab("") +
  ylab("Peptide number") +
  theme_bw()
```

## Human keratins

```{r}
# Load contaminant list
cont <- read_delim("identified_contaminants.txt", 
    delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(str_detect(protein_id, "HUMAN"))
cont_list <- paste(cont$id, collapse = "|")

# Load dataset
human <- read_delim("proteinGroups.txt", 
    delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(str_detect(protein_i_ds, cont_list)) %>%
  filter(!str_detect(protein_i_ds, "CANLF|HORSE")) %>%
  select(protein_i_ds, starts_with("razor_unique_peptides_")) 

# Total number of proteins and peptides
human_total <- human %>%
  pivot_longer(names_to = "stagetip", values_to = "peptides", 2:23) %>%
  mutate(stagetip = gsub('razor_unique_peptides_', '', stagetip)) %>%
  filter(peptides > 0) %>%
  group_by(stagetip) %>%
  summarize(protein_count = n(), 
            peptide_count = sum(peptides))

# Add metadata
human_total$stagetip <- toupper(human_total$stagetip)
human_total <- left_join(human_total, meta)

# Summarize
human_summary <- human_total %>%
  group_by(method) %>%
  summarize(mean_protein = mean(protein_count), sd_protein = sd(protein_count), 
            mean_peptide = mean(peptide_count), sd_peptide = sd(peptide_count))

# Change labels
human_summary <- human_summary %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))

# Reorder for plotting
human_summary$method <- factor(human_summary$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot (not included in the manuscript)
human_proteins <- ggplot(human_summary, aes(x=method, y=mean_protein, fill=method)) + 
  geom_bar(stat="identity", position ="dodge") +
  geom_errorbar(aes(ymin=mean_protein-sd_protein, ymax=mean_protein+sd_protein), 
                width=0.2, position=position_dodge(0.9)) +
  scale_fill_manual(values=met.brewer("Cassatt2", 7), guide="none") +
  xlab("") +
  ylab("Protein number") +
  theme_bw()

human_peptides <- ggplot(human_summary, aes(x=method, y=mean_peptide, fill=method)) + 
  geom_bar(stat="identity", position ="dodge") +
  geom_errorbar(aes(ymin=mean_peptide-sd_peptide, ymax=mean_peptide+sd_peptide), 
                width=0.2, position=position_dodge(0.9)) +
  scale_fill_manual(values=met.brewer("Cassatt2", 7), guide="none") +
  xlab("") +
  ylab("Peptide number") +
  theme_bw()
```

### Lactotransferrin

```{r}
# Peptide count
lt <- read_delim("equus_canis_maxquant/proteinGroups.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(str_detect(fasta_headers, "Lactotransferrin")) %>%
  select(starts_with("razor_unique_peptides_")) %>%
  pivot_longer(values_to="peptides", names_to="stagetip", 1:22) %>%
  mutate(stagetip = gsub('razor_unique_peptides_', '', stagetip))

# Add metadata
lt$stagetip <- toupper(lt$stagetip)
lt <- left_join(lt, meta) %>%
  filter(!sample == "blank")

# Summarize
lt_summary <- lt %>%
  group_by(method) %>%
  summarize(mean_peptides = mean(peptides), sd_peptides = sd(peptides))

# Change labels
lt_summary <- lt_summary %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))

# Reorder for plotting
lt_summary$method <- factor(lt_summary$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot
lacto_peptides <- ggplot(lt_summary, aes(x=method, y=mean_peptides)) + 
  geom_bar(stat="identity", position ="dodge", fill= "#b695bc") +
  geom_errorbar(aes(ymin=mean_peptides-sd_peptides, ymax=mean_peptides+sd_peptides), 
                width=0.2, position=position_dodge(0.9)) +
  xlab("") +
  ylab("Peptides") +
  theme_bw()

# Sequence coverage
lt <- read_delim("proteinGroups.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(str_detect(fasta_headers, "Lactotransferrin")) %>%
  select(starts_with("sequence_coverage_")) %>%
  select(-sequence_coverage_percent) %>%
  pivot_longer(values_to="coverage", names_to="stagetip", 1:22) %>%
  mutate(stagetip = gsub('sequence_coverage_', '', stagetip)) %>%
  mutate(stagetip = gsub('_percent', '', stagetip))

# Add metadata
lt$stagetip <- toupper(lt$stagetip)
lt <- left_join(lt, meta) %>%
  filter(!sample == "blank")

# Summarize
lt_summary <- lt %>%
  group_by(method) %>%
  summarize(mean_coverage = mean(coverage), sd_coverage = sd(coverage))

# Change labels
lt_summary <- lt_summary %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))

# Reorder for plotting
lt_summary$method <- factor(lt_summary$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot (Figure 4A-B)
lacto_coverage <- ggplot(lt_summary, aes(x=method, y=mean_coverage)) + 
  geom_bar(stat="identity", position ="dodge", fill= "#b695bc") +
  geom_errorbar(aes(ymin=mean_coverage-sd_coverage, ymax=mean_coverage+sd_coverage), 
                width=0.2, position=position_dodge(0.9)) +
  xlab("") +
  ylab("Sequence coverage (%)") +
  theme_bw()
```

## Protein coverage

Change protein IDs according to what you want to look at.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load peptide files
peptide <- read_delim("equus_canis_maxquant/peptides.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(str_detect(proteins, "A0A9L0TS53|F6WD70")) %>%
  dplyr::select(starts_with("lfq_intensity_p"), start_position, end_position)

# Expand numbers between start and end to get a dataframe of all covered positions
peptide <- peptide %>%
  mutate(covered = map2(start_position, end_position, seq, by=1))  
pos_list <- peptide %>% dplyr::select(covered) %>%
  mutate(covered = map(covered, ~as_data_frame(t(.x)))) %>% unnest()
peptide <- cbind(peptide, pos_list)

# Gather to long form
peptide <- peptide %>%
  gather("remove", "position", 26:50) %>% 
  dplyr::select(-start_position, -end_position, -covered, -remove) %>%
  filter(position > 0) %>%
  gather("stagetip", "intensity", 1:22) %>%
  filter(intensity > 0) %>%
  separate(stagetip, c("remove", "remove2", "stagetip"), sep="_") %>%
  select(-remove, -remove2)

# Sum intensity per position
coverage_summary <- peptide %>%
  group_by(position, stagetip) %>%
  summarize(total_intensity = sum(intensity))

# Add metadata, remove blank
coverage_summary$stagetip <- toupper(coverage_summary$stagetip)
coverage_summary <- left_join(coverage_summary, meta) %>%
  filter(!sample == "blank")

## Count number of covered amino acids
coverage_count <- coverage_summary %>%
  group_by(method, sample) %>%
  summarize(covered_aa = n())

# Summarize even more
coverage_count_summary <- coverage_count %>%
  group_by(method) %>%
  summarize(mean_aa = mean(covered_aa),
            sd_aa = sd(covered_aa))

# ANOVA
aov_test <- aov(covered_aa ~ method, data = coverage_count)

# Summary of the analysis
summary(aov_test)

# Post-hoc test
TukeyHSD(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(covered_aa ~ method, data = coverage_count)

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals)

# log2 intensity
coverage_summary$intensity_log2 <- log2(coverage_summary$total_intensity)

# Average by method
coverage_summary <- coverage_summary %>%
  group_by(method, position) %>%
  summarize(mean_intensity = mean(intensity_log2))

# Change labels
coverage_summary <- coverage_summary %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))

# Reorder for plotting
coverage_summary$method <- factor(coverage_summary$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot (leads to Figure 5)
ggplot(coverage_summary, aes(x=position, y=method)) +
  geom_raster(aes(fill=mean_intensity)) +
  scale_fill_gradient(low="#e9eaf2ff", high="#574571", name="Mean intensity (log2)",
                      limits=c(15, 35)) +
  scale_y_discrete(limits = rev) +
  ylab("") +
  xlab("Amino acid position") +
  theme_bw()
```

## PTMs

```{r}
# Load evidence file and clean, filter for horse-only peptides
evidence <- read_delim("equus_canis_maxquant/evidence.txt", delim = "\t") %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(intensity > 0) %>%
  filter(!str_detect(proteins, "CANLF")) %>%
  filter(str_detect(proteins, "HORSE")) %>%
  filter(!str_detect(proteins, "CON__P13647|F6Z8D6|P08730-1|P02538|F6Q1M4|A0A3Q2KP58|P19013|F6S8W6|P50446|F7D5X8|Q5XQN5"))
```

### Oxidation of M

```{r}
# Calculate 
oxidation_m <- evidence %>%
  mutate(count_m = str_count(sequence, "M"),
         count_oxi_m = str_count(modified_sequence, "M\\("),
         fraction_oxi_m = count_oxi_m / count_m) %>%
  filter(count_m > 0) %>%
  group_by(experiment) %>%
  summarise(mean_oxi_m = weighted.mean(fraction_oxi_m, intensity, na.rm=TRUE),
            number_peptides = n())

# Add metadata
oxidation_m <- left_join(oxidation_m, meta, by=c("experiment" = "stagetip"))

# ANOVA
aov_test <- aov(mean_oxi_m ~ method, data = oxidation_m)

# Summary of the analysis
summary(aov_test)

# Post-hoc test
TukeyHSD(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(mean_oxi_m ~ method, data = oxidation_m_stats)

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals)

# Summarize
oxi_summary <- oxidation_m %>%
  group_by(method) %>%
  summarize(mean_oxi = mean(mean_oxi_m), sd_oxi = sd(mean_oxi_m),
            mean_peptides = mean(number_peptides))

# Change labels
oxi_summary <- oxi_summary %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))

# Reorder for plotting
oxi_summary$method <- factor(oxi_summary$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot (not included in manuscript)
oxi_plot <- ggplot(oxi_summary, aes(x=method, y=mean_oxi)) + 
  geom_bar(stat="identity", position ="dodge", fill=met.brewer("Cassatt2", 1)) +
  geom_errorbar(aes(ymin=mean_oxi-sd_oxi, ymax=mean_oxi+sd_oxi), 
                width=0.2, position=position_dodge(0.9)) +
  xlab("") +
  ylab("Oxidation M") +
  theme_bw()
```

### Carbamidomethylation of C

```{r}
# Calculate 
carb <- evidence %>%
  mutate(count_c = str_count(sequence, "C"),
         count_carb = str_count(modified_sequence, "C\\("),
         fraction_carb = count_carb / count_c) %>%
  filter(count_c > 0) %>%
  group_by(experiment) %>%
  summarise(mean_carb = weighted.mean(fraction_carb, intensity, na.rm=TRUE),
            number_peptides = n())

# Add metadata
carb <- left_join(carb, meta, by=c("experiment" = "stagetip"))

# ANOVA
aov_test <- aov(mean_carb ~ method, data = carb)

# Summary of the analysis
summary(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(mean_carb ~ method, data = carb)

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals)

# Summarize
carb_summary <- carb %>%
  group_by(method) %>%
  dplyr::summarize(mean_carbo = mean(mean_carb), sd_carb = sd(mean_carb))

# Change labels
carb_summary <- carb_summary %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))

# Reorder for plotting
carb_summary$method <- factor(carb_summary$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot (not included in manuscript)
carb_plot <- ggplot(carb_summary, aes(x=method, y=mean_carbo)) + 
  geom_bar(stat="identity", position ="dodge", fill=met.brewer("Cassatt2", 1)) +
  geom_errorbar(aes(ymin=mean_carbo-sd_carb, ymax=mean_carbo+sd_carb), 
                width=0.2, position=position_dodge(0.9)) +
  xlab("") +
  ylab("Carbamidomethylation C") +
  theme_bw()
```

### Pyro-Glu

```{r}
# Calculate 
pyro <- evidence %>%
  mutate(count_e = str_count(sequence, "E"),
         count_pyro = str_count(modified_sequence, "\\(Glu\\-"),
         fraction_pyro = count_pyro / count_e) %>%
  filter(count_e > 0) %>%
  group_by(experiment) %>%
  summarise(mean_pyro = weighted.mean(fraction_pyro, intensity, na.rm=TRUE),
            number_peptides = n())

# Add metadata
pyro <- left_join(pyro, meta, by=c("experiment" = "stagetip"))

# ANOVA
aov_test <- aov(mean_pyro ~ method, data = pyro)

# Summary of the analysis
summary(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(mean_pyro ~ method, data = pyro)

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals)
```

### Hydroxyproline

```{r}
# Calculate 
hydro <- evidence %>%
  mutate(count_p = str_count(sequence, "P"),
         count_hydro = str_count(modified_sequence, "P\\("),
         fraction_hydro = count_hydro / count_p) %>%
  filter(count_p > 0) %>%
  group_by(experiment) %>%
  summarise(mean_hydro = weighted.mean(fraction_hydro, intensity, na.rm=TRUE),
            number_peptides = n())

# Add metadata
hydro <- left_join(hydro, meta, by=c("experiment" = "stagetip"))

# ANOVA
aov_test <- aov(mean_hydro ~ method, data = hydro)

# Summary of the analysis
summary(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(mean_hydro ~ method, data = hydro)

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals)

# Summarize
hydro_summary <- hydro %>%
  group_by(method) %>%
  dplyr::summarize(mean_hydrox = mean(mean_hydro), sd_hydro = sd(mean_hydro))

# Change labels
hydro_summary <- hydro_summary %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))

# Reorder for plotting
hydro_summary$method <- factor(hydro_summary$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot (Figure S2C)
hydro_plot <- ggplot(hydro_summary, aes(x=method, y=mean_hydrox)) + 
  geom_bar(stat="identity", position ="dodge", fill=met.brewer("Cassatt2", 1)) +
  geom_errorbar(aes(ymin=mean_hydrox-sd_hydro, ymax=mean_hydrox+sd_hydro), 
                width=0.2, position=position_dodge(0.9)) +
  xlab("") +
  ylab("Hydroxyproline") +
  theme_bw()
```

### Oxidation (WH)

Note that UV-induced modifications are from a different MaxQuant run! It's also in the PRIDE repository.

```{r}
# Load evidence file and clean, filter for horse-only peptides
evidence_uv <- read_delim("equus_uv_maxquant/evidence.txt", delim = "\t") %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(intensity > 0) %>%
  filter(str_detect(proteins, "HORSE")) %>%
  filter(!str_detect(proteins, "CON__P13647|F6Z8D6|P08730-1|P02538|F6Q1M4|A0A3Q2KP58|P19013|F6S8W6|P50446|F7D5X8|Q5XQN5"))

# Calculate 
wh <- evidence_uv %>%
  mutate(count_wh = str_count(sequence, "W|H"),
         count_oxiwh = str_count(modified_sequence, "\\(WH\\)"),
         fraction_oxiwh = count_oxiwh / count_wh) %>%
  filter(count_wh > 0) %>%
  group_by(experiment) %>%
  summarise(mean_oxiwh = weighted.mean(fraction_oxiwh, intensity, na.rm=TRUE),
            number_peptides = n())

# Add metadata
wh <- left_join(wh, meta, by=c("experiment" = "stagetip"))

# ANOVA
aov_test <- aov(mean_oxiwh ~ method, data = wh)

# Summary of the analysis
summary(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(mean_oxiwh ~ method, data = wh)

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals)

# Summarize
wh_summary <- wh %>%
  group_by(method) %>%
  summarize(mean_wh = mean(mean_oxiwh), sd_wh = sd(mean_oxiwh),
            mean_peptides = mean(number_peptides))

# Change labels
wh_summary <- wh_summary %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))

# Reorder for plotting
wh_summary$method <- factor(wh_summary$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot (Figure S2B)
wh_plot1 <- ggplot(wh_summary, aes(x=method, y=mean_wh)) + 
  geom_bar(stat="identity", position ="dodge", fill=met.brewer("Cassatt2", 1)) +
  geom_errorbar(aes(ymin=mean_wh-sd_wh, ymax=mean_wh+sd_wh), 
                width=0.2, position=position_dodge(0.9)) +
  xlab("") +
  ylab("Oxidation WH") +
  theme_bw()
```

### Dioxidation (WH)

```{r}
# Calculate 
wh <- evidence_uv %>%
  mutate(count_wh = str_count(sequence, "W|H"),
         count_oxiwh = str_count(modified_sequence, "\\(Di-oxi"),
         fraction_oxiwh = count_oxiwh / count_wh) %>%
  filter(count_wh > 0) %>%
  group_by(experiment) %>%
  summarise(mean_oxiwh = weighted.mean(fraction_oxiwh, intensity, na.rm=TRUE),
            number_peptides = n())

# Add metadata
wh <- left_join(wh, meta, by=c("experiment" = "stagetip"))

# ANOVA
aov_test <- aov(mean_oxiwh ~ method, data = wh)

# Summary of the analysis
summary(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(mean_oxiwh ~ method, data = wh)

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals)

# Summarize
wh_summary <- wh %>%
  group_by(method) %>%
  summarize(mean_wh = mean(mean_oxiwh), sd_wh = sd(mean_oxiwh),
            mean_peptides = mean(number_peptides))

# Change labels
wh_summary <- wh_summary %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))

# Reorder for plotting
wh_summary$method <- factor(wh_summary$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot (Figure S2D)
wh_plot2 <- ggplot(wh_summary, aes(x=method, y=mean_wh)) + 
  geom_bar(stat="identity", position ="dodge", fill=met.brewer("Cassatt2", 1)) +
  geom_errorbar(aes(ymin=mean_wh-sd_wh, ymax=mean_wh+sd_wh), 
                width=0.2, position=position_dodge(0.9)) +
  xlab("") +
  ylab("Dioxidation WH") +
  theme_bw()
```

## Deamidation

Here, you first need to calculate deamidation from the evidence file using the script published in:

Mackie, M., Rüther, P., Samodova, D., Di Gianvincenzo, F., Granzotto, C., Lyon, D., Peggie, D. A., Howard, H., Harrison, L., Jensen, L. J., Olsen, J. V., & Cappellini, E. (2018). Palaeoproteomic Profiling of Conservation Layers on a 14th Century Italian Wall Painting. Angewandte Chemie, 57(25), 7369–7374.

```{r}
# Load evidence file and clean, filter for horse-only peptides
evidence <- read_delim("equus_canis_maxquant/evidence.txt", delim = "\t") %>%
  filter(!Reverse %in% c("+")) %>%
  filter(Intensity > 0) %>%
  filter(!str_detect(Proteins, "CANLF")) %>%
  filter(str_detect(Proteins, "HORSE")) %>%
  filter(!str_detect(Proteins, "CON__P13647|F6Z8D6|P08730-1|P02538|F6Q1M4|A0A3Q2KP58|P19013|F6S8W6|P50446|F7D5X8|Q5XQN5"))

# Load deamidation output
deam <- read.csv("Deamidation.txt", sep="") %>%
  clean_names()

# Extract stagetip number
deam <- deam %>%
  separate(raw_file, c("remove", "stagetip"), sep="DDA_") %>%
  separate(stagetip, c("stagetip", "remove1"), sep="_") %>%
  select(-remove, -remove1)

# Add metadata
deam <- left_join(deam, meta)

# Separate into N and Q for stats
deam_n <- deam %>% filter(n_q == "N")
deam_q <- deam %>% filter(n_q == "Q")

### Stats N

# ANOVA
aov_test <- aov(mean ~ method, data = deam_n)

# Summary of the analysis
summary(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(mean ~ method, data = deam_n)

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals)

### Stats Q

# ANOVA
aov_test <- aov(mean ~ method, data = deam_q)

# Summary of the analysis
summary(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(mean ~ method, data = deam_q)

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals)

### Plotting

# Summarize
deam_summary <- deam %>%
  group_by(method, n_q) %>%
  dplyr::summarize(mean_deam = mean(mean), sd_deam = sd(mean))

# Change labels
deam_summary <- deam_summary %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))

# Reorder for plotting
deam_summary$method <- factor(deam_summary$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot (Figure S2E)
deam_plot <- ggplot(deam_summary, aes(x=method, y=mean_deam)) + 
  geom_bar(stat="identity", position ="dodge", fill=met.brewer("Cassatt2", 1)) +
  geom_errorbar(aes(ymin=mean_deam-sd_deam, ymax=mean_deam+sd_deam), 
                width=0.2, position=position_dodge(0.9)) +
  facet_wrap(~n_q) +
  xlab("") +
  ylab("Deamidation") +
  theme_bw()
```

## Peptide length 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load evidence file and clean
evidence <- read_delim("equus_canis_maxquant/evidence.txt", delim = "\t") %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(intensity > 0) %>%
  filter(!str_detect(proteins, "CANLF")) %>%
  filter(str_detect(proteins, "HORSE")) %>%
  filter(!str_detect(proteins, "CON__P13647|F6Z8D6|P08730-1|P02538|F6Q1M4|A0A3Q2KP58|P19013|F6S8W6|P50446|F7D5X8|Q5XQN5"))

# Add metadata
evidence <- left_join(evidence, meta, by=c("experiment"="stagetip"))

# Summarize
length_summary <- evidence %>%
  group_by(experiment, method) %>%
  dplyr::summarise(mean_length = weighted.mean(x=length, w=intensity, na.rm=TRUE))

# ANOVA
aov_test <- aov(mean_length ~ method, data = length_summary)

# Summary of the analysis
summary(aov_test)

# Post-hoc test
TukeyHSD(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(mean_length ~ method, data = length_summary)

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals)

# Summarize
length_plot <- length_summary %>%
  group_by(method) %>%
  summarize(mean_lengths = mean(mean_length), sd_lengths = sd(mean_length))

# Change labels
length_plot <- length_plot %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))

# Reorder for plotting
length_plot$method <- factor(length_plot$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot (leads to Figure S2A)
length_fig <- ggplot(length_plot, aes(x=method, y=mean_lengths)) + 
  geom_bar(stat="identity", position ="dodge", fill=met.brewer("Cassatt2", 1)) +
  geom_errorbar(aes(ymin=mean_lengths-sd_lengths, ymax=mean_lengths+sd_lengths), 
                width=0.2, position=position_dodge(0.9)) +
  ylim(0,27) +
  xlab("") +
  ylab("Peptide length") +
  theme_bw()
```

## Horse protein intensity

```{r}
# Load dataset
intensity <- read_delim("equus_canis_maxquant/evidence.txt", 
    delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(!str_detect(proteins, "CANLF")) %>%
  filter(str_detect(proteins, "HORSE")) %>%
  filter(!str_detect(proteins, "CON__P13647|F6Z8D6|P08730-1|P02538|F6Q1M4|A0A3Q2KP58|P19013|F6S8W6|P50446|F7D5X8|Q5XQN5")) %>%
  filter(intensity > 0) %>%
  select(proteins, experiment, intensity)

# Summarize
intensity <- intensity %>%
  group_by(experiment) %>%
  summarize(total_intensity = sum(intensity))

# Add metadata
meta <- meta %>%
  filter(weight_mg > 0) %>%
  mutate(scale = max(weight_mg)/weight_mg)
intensity <- left_join(intensity, meta, by=c("experiment"="stagetip")) %>%
  filter(!sample == "blank")

# Normalize by weight
intensity$int_scaled <- intensity$total_intensity * intensity$scale

# ANOVA
aov_test <- aov(int_scaled ~ method, data = intensity)

# Summary of the analysis
summary(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(total_intensity ~ method, data = intensity_sum)

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals)

# Summarize
intensity_plot <- intensity %>%
  group_by(method) %>%
  summarize(mean_intensity = mean(int_scaled), sd_intensity = sd(int_scaled))

# Change labels
intensity_plot <- intensity_plot %>%
    mutate(method = gsub('no_contamination', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('uv', 'UV', method))  %>%
    mutate(method = gsub('no_cleaning', 'No decont.', method))

# Reorder for plotting
intensity_plot$method <- factor(intensity_plot$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Plot (not included in manuscript)
intensity_fig <- ggplot(intensity_plot, aes(x=method, y=mean_intensity)) + 
  geom_bar(stat="identity", position ="dodge", fill=met.brewer("Cassatt2", 1)) +
  geom_errorbar(aes(ymin=mean_intensity-sd_intensity,
                    ymax=mean_intensity+sd_intensity), 
                width=0.2, position=position_dodge(0.9)) +
  xlab("") +
  ylab("Horse intensity") +
  theme_bw()
```

## Overall PCA

```{r echo=FALSE, message=FALSE, warning=FALSE}
gene <- read_delim("equus_canis_maxquant/proteinGroups.txt", 
    delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  #filter(str_detect(protein_i_ds, "CANLF|HORSE")) %>%
  select(protein_i_ds, starts_with("lfq_")) %>%
  select(-lfq_intensity_p0408)

# Filter for presence in two samples 
gene$presence <- rowSums(gene[2:22] != 0)
gene <- gene %>%
  filter(presence > 1) %>%
  dplyr::select(-presence)

# Transpose
gene <- gene %>%
  gather(stagetip, intensity, 2:22) %>%
  group_by(stagetip, protein_i_ds) %>%
  summarize(intensity = sum(intensity)) %>%
  ungroup() %>%
  spread(protein_i_ds, intensity) %>%
  replace(is.na(.), 0)

# Fix sample names
gene <- gene %>%
  mutate(stagetip = gsub('lfq_intensity_', '', stagetip))
gene$stagetip <- toupper(gene$stagetip)

# Pseudocount zero replacement
gene[,2:ncol(gene)] <- gene[,2:ncol(gene)]+1

# Log-transformation
gene[,2:ncol(gene)] <- log2(gene[,2:ncol(gene)])

# Create PCA object
pca_all <- prcomp(gene[, 2:133], scale=FALSE)

# Extract values
pca_out <- as_tibble(pca_all$x)

# Add metadata
pca_out$stagetip <- gene$stagetip
pca_out <- left_join(pca_out, meta)

# Calculate percent explained per PC
percentage <- round(pca_all$sdev^2 / sum(pca_all$sdev^2) * 100, 2)
percentage <- paste(colnames(pca_out), "(", paste(
  as.character(percentage), "%", ")", sep="") )

# Get variables with biggest impact for PC1 and PC2
pca_all$rotation[,1] %>% sort() %>% head(1)
pca_all$rotation[,1] %>% sort(decreasing = TRUE) %>% head(1)
pca_all$rotation[,2] %>% sort() %>% head(1)
pca_all$rotation[,2] %>% sort(decreasing = TRUE) %>% head(1)

# Get rownames for each protein
which(rownames(pca_all$rotation) %in% 
        c("tr|A0A8I3P9H6|A0A8I3P9H6_CANLF;tr|F6XL78|F6XL78_HORSE",
          "tr|A0A8I3MKE4|A0A8I3MKE4_CANLF",
          "tr|A0A8I3SBH3|A0A8I3SBH3_CANLF;sp|Q865C0|CAH6_CANLF",
          "tr|A0A8I3P0I4|A0A8I3P0I4_CANLF"))

# Create matrix of x (PC1) coordinates and y (PC2)
l.x <- cbind(pca_all$rotation[,1][c(46,77,84,115)]) * 150
l.y <- cbind(pca_all$rotation[,2][c(46,77,84,115)]) * 150

# Plot (leads to Figure 2B)
pca <- ggplot(data = pca_out, aes(x = PC1, y = PC2, color=method)) +
  geom_point(size=6, stroke=1) +
  xlab(percentage[1]) + 
  ylab(percentage[2]) +
  scale_color_manual(values=met.brewer(n=7, name="Cassatt2")) +
  theme_minimal(base_size = 10) +
  annotate("segment", x=0, xend=l.x, y=0, yend=l.y, colour="black", 
           size=1, arrow=arrow()) + 
  annotate("text", x=l.x, y=l.y, label=rownames(l.x), color="black")

# Basic permanova with euclidean distance as metric
library(vegan)
pca_meta <- left_join(gene, meta)
permanova <- adonis(gene[, 2:133] ~ method, 
                    data=pca_meta,
                    permutations=999,
                    method="euclidean")
permanova

# Checking assumptions
dist <- vegdist(gene[, 2:133])
anova(betadisper(dist, pca_meta$method))
```

## Horse proteins

From MaxQuant run without dog proteome, also included in PRIDE submission.

```{r}
# Load datasets and clean
horse <- read_delim("equus_only_maxquant/proteinGroups.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names()  %>% 
  filter(!reverse %in% c("+"))  %>% 
  filter(!str_detect(protein_i_ds, "CON_")) %>%
  dplyr::select(fasta_headers, 33:35, 51:53)

# Check number of proteins
horse <- horse %>%
  pivot_longer(names_to = "stagetip", values_to = "peptides", 2:7) %>%
  mutate(stagetip = gsub('razor_unique_peptides_', '', stagetip)) %>%
  filter(peptides > 0) %>%
  group_by(stagetip) %>%
  summarize(protein_count = n())

# Add metadata
horse$stagetip <- toupper(horse$stagetip)
horse <- left_join(horse, meta)

# Summarize
horse_summary <- horse %>%
  group_by(method) %>%
  summarize(mean_protein = mean(protein_count), sd_protein = sd(protein_count))

# Same for PSMs
psms <- read_delim("/path_to_wd/evidence.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names()  %>% 
  filter(!reverse %in% c("+")) %>%
  filter(!str_detect(proteins, "CON_")) %>%
  group_by(experiment) %>%
  summarize(psm_nr = n())

# Add metadata
psms <- left_join(psms, meta, by=c("experiment" = "stagetip")) %>%
  filter(method %in% c("no_cleaning", "no_contamination"))

# Summarize
psm_summary <- psms %>%
  group_by(method) %>%
  summarize(mean_psm = mean(psm_nr), sd_psm = sd(psm_nr))
```

## Horse PCA

From MaxQuant run without dog proteome, also included in PRIDE submission.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load datasets and clean
gene_horse <- read_delim("equus_only_maxquant/proteinGroups.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names()  %>% 
  filter(!reverse %in% c("+"))  %>% 
  filter(str_detect(fasta_headers, "HORSE")) %>%
  filter(!str_detect(fasta_headers, "Keratin")) %>%
  dplyr::select(fasta_headers, 130:132, 148:150)

# Extract protein names
gene_horse <- gene_horse %>%
  separate(fasta_headers, c("protein", "remove"), " ") %>%
  dplyr::select(-remove)

# Filter for presence in two samples 
gene_horse$presence <- rowSums(gene_horse[2:7] != 0)
gene_horse <- gene_horse %>%
  filter(presence > 1) %>%
  dplyr::select(-presence)

# Transpose
gene_horse <- gene_horse %>%
  gather(stagetip, intensity, 2:7) %>%
  group_by(stagetip, protein) %>%
  summarize(intensity = sum(intensity)) %>%
  ungroup() %>%
  spread(protein, intensity) %>%
  replace(is.na(.), 0)

# Fix sample names
gene_horse <- gene_horse %>%
  mutate(stagetip = gsub('lfq_intensity_', '', stagetip))
gene_horse$stagetip <- toupper(gene_horse$stagetip)

# Pseudocount zero replacement
gene_horse[,2:ncol(gene_horse)] <- gene_horse[,2:ncol(gene_horse)]+1

# Log-transformation
gene_horse[,2:ncol(gene_horse)] <- log2(gene_horse[,2:ncol(gene_horse)])

# Create PCA object
pca_all <- prcomp(gene_horse[, 2:36], scale=FALSE)

# Extract values
pca_out <- as_tibble(pca_all$x)

# Add metadata
pca_out$stagetip <- gene_horse$stagetip
pca_out <- left_join(pca_out, meta)

# Calculate percent explained per PC
percentage <- round(pca_all$sdev^2 / sum(pca_all$sdev^2) * 100, 2)
percentage <- paste(colnames(pca_out), "(", paste(
  as.character(percentage), "%", ")", sep="") )

# Get variables with biggest impact for PC1 and PC2
pca_all$rotation[,1] %>% sort() %>% head(1)
pca_all$rotation[,1] %>% sort(decreasing = TRUE) %>% head(1)
pca_all$rotation[,2] %>% sort() %>% head(1)
pca_all$rotation[,2] %>% sort(decreasing = TRUE) %>% head(1)

# Get rownames for each protein
which(rownames(pca_all$rotation) %in% 
        c("tr|F6XL78|F6XL78_HORSE", "tr|A0A3Q2H9I4|A0A3Q2H9I4_HORSE",
          "tr|F7BCM8|F7BCM8_HORSE", "tr|F7BND9|F7BND9_HORSE"))

# Create matrix of x (PC1) coordinates and y (PC2) (duplicates removed)
l.x <- cbind(pca_all$rotation[,1][c(9,24,28,32)]) * 100
l.y <- cbind(pca_all$rotation[,2][c(9,24,28,32)]) * 100

# Plot (leads to Figure 2A)
horse_pca <- ggplot(data = pca_out, aes(x = PC1, y = PC2, color=method)) +
  geom_point(size=6, stroke=1) +
  xlab(percentage[1]) + 
  ylab(percentage[2]) +
  scale_color_manual(values=met.brewer(n=4, name="Cassatt2")) +
  theme_minimal(base_size = 10) +
  annotate("segment", x=0, xend=l.x, y=0, yend=l.y, colour="black", 
           size=1, arrow=arrow()) + 
  annotate("text", x=l.x, y=l.y, label=rownames(l.x), color="black")
```

## Identified MS2 spectra

From MaxQuant run without dog proteome, also included in PRIDE submission.

```{r}
# Load summary file
summary <- read_delim("equus_only_maxquant/summary.txt", 
    delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  janitor::clean_names() %>%
  filter(!is.na(experiment)) %>%
  filter(!experiment == "P0408") %>%
  filter(!raw_file == "20230525_EXPL8_nLC11_GT_collab_77min_DDA_P0389")

# Add metadata
summary <- left_join(summary, meta, by=c("experiment"="stagetip")) %>%
  filter(method %in% c("no_contamination", "no_cleaning"))

# ANOVA
aov_test <- aov(ms_ms_identified ~ method, data = summary)

# Summary of the analysis
summary(aov_test) 

# Post-hoc test
TukeyHSD(aov_test)

# Homogeneity of variances
plot(aov_test, 1)
leveneTest(ms_ms_identified ~ method, data = summary) # all good

# Normality
plot(aov_test, 2)
aov_residuals <- residuals(object = aov_test)
shapiro.test(x = aov_residuals) # all good

# Summarize
mean_summary <- summary %>%
  group_by(method) %>%
  summarize(mean_id_percent_msms = mean(ms_ms_identified_percent),
            sd_id_percent_msms = sd(ms_ms_identified_percent),
            mean_msms = mean(ms_ms), sd_msms = sd(ms_ms),
            mean_id_msms = mean(ms_ms_identified),
            sd_id_msms = sd(ms_ms_identified))

# Change labels
mean_summary <- mean_summary %>%
    mutate(method = gsub('original', 'Original', method)) %>%
    mutate(method = gsub('bleach', 'Bleach', method)) %>%
    mutate(method = gsub('edta', 'EDTA', method)) %>%
    mutate(method = gsub('hcl', 'HCl', method)) %>%
    mutate(method = gsub('uv', 'UV', method)) %>%
    mutate(method = gsub('water', 'Water', method)) %>%
    mutate(method = gsub('no_decont', 'No decont.', method))

# Reorder for plotting
mean_summary$method <- factor(mean_summary$method, 
                          levels=c("Original", "Bleach", "EDTA","HCl",
                                   "Water", "UV", "No decont."))

# Figure for the number of MS2 spectra (not included in manuscript)
ggplot(mean_summary, aes(x=method, y=mean_msms, fill=method)) +
  geom_bar(stat="identity") +
  geom_errorbar(aes(ymin=mean_msms-sd_msms, ymax=mean_msms+sd_msms), 
                width=0.2, position=position_dodge(0.9)) +
  scale_fill_manual(values=met.brewer("Hiroshige", 8), guide="none") +
  xlab("Method") +
  ylab("MS2 spectra") +
  theme_bw()

# Figure for the number of identified MS2 spectra (not included in manuscript)
ggplot(mean_summary, aes(x=method, y=mean_identified_msms, fill=method)) +
  geom_bar(stat="identity") +
  geom_errorbar(aes(ymin=mean_identified_msms-sd_identified_msms, ymax=mean_identified_msms+sd_identified_msms), 
                width=0.2, position=position_dodge(0.9)) +
  scale_fill_manual(values=met.brewer("Hiroshige", 8), guide="none") +
  xlab("Method") +
  ylab("Identified MS2 spectra") +
  theme_bw()

# Figure for the percentage of identified MS2 spectra (not included in manuscript)
ggplot(mean_summary, aes(x=method, y=mean_id_percent_msms, fill=method)) +
  geom_bar(stat="identity") +
  geom_errorbar(aes(ymin=mean_id_percent_msms-sd_id_percent_msms, ymax=mean_id_percent_msms+sd_id_percent_msms), 
                width=0.2, position=position_dodge(0.9)) +
  scale_fill_manual(values=met.brewer("Hiroshige", 8), guide="none") +
  xlab("Method") +
  ylab("Identified MS2 spectra %") +
  theme_bw()

# Figure for the identified peptides (not included in manuscript)
ggplot(mean_summary, aes(x=method, y=mean_peptides, fill=method)) +
  geom_bar(stat="identity") +
  geom_errorbar(aes(ymin=mean_peptides-sd_peptides, ymax=mean_peptides+sd_peptides), 
                width=0.2, position=position_dodge(0.9)) +
  scale_fill_manual(values=met.brewer("Hiroshige", 8), guide="none") +
  xlab("Method") +
  ylab("Identified peptides") +
  theme_bw()
```
